\subsection{Challenges for future research} \label{sec:recommendations}

Viewing the concepts in Section \ref{sec:by-example} through the
practice scenarios we have discussed, we can describe the following
challenges for research in computational serendipity.

\begin{itemize}
\item \textbf{Autonomy}: Our case studies in Section
  \ref{sec:computational-serendipity} highlight the potential value of
  increased autonomy on the system side.
%% The thought experiment in Section
%% \ref{sec:ww} develops a design illustrating the relationship between
%% creativity at the level of artefacts (e.g.~new poems) and
%% creativity at the level of \emph{problem specification} (learning
%% new poetic concepts).
The search for connections that make raw data into ``strategic data''
is an appropriate theme for research in computational intelligence and
machine learning to grapple with.  In the standard cybernetic model,
we control computers, and we also control the computer's operating
context.  There is little room for serendipity if there is nothing
outside of our direct control.  In contrast with the mainstream model,
von Foerster \citeyear[p. 286]{von2003cybernetics} advocated a
second-order cybernetics in which ``the observer who enters the system
shall be allowed to stipulate his own purpose.''  Accordingly, \emph{a
  primary challenge to the serendipitous operation of computers is
  developing computational agents that specify their own problems.}
\end{itemize}

\begin{itemize}
\item \textbf{Learning}: Each of the case studies considered in
  Section \ref{sec:computational-serendipity} is able to learn from
  experience.  As we considered ways to enhance measures of
  serendipity in these examples, we were led to consider computational
  agents that participate meaningfully in ``our world'' rather than in
  a circumscribed microdomain.  \emph{A second challenge is for
    computational agents to learn more and more about the world we
    live in.}
\end{itemize}

\begin{itemize}
\item \textbf{Sociality}: We may be aided in our pursuit of the
  ``smart mind'' required for serendipity by recalling Turing's
  proposal that computers should ``be able to converse with each other
  to sharpen their wits'' \cite{turing-intelligent}.  Turing
  recognised that computers would have to be coached in the direction
  of social learning, but that once they attain that standard they
  will learn much more quickly.  The four supportive factors for
  serendipity described in this paper resemble nothing more than our
  experience of social reality.  \emph{A third challenge is for
    computational agents to interact in a recognisably social way with
    us and with each other, resulting in emergent effects.}
\end{itemize}

\begin{itemize}
\item \textbf{Embedded evaluation}:
  \citeA{stakeholder-groups-bookchapter} outline a general programme
  for computational creativity, and examined perceptions of creativity
  in computational systems found among members of the general public,
  Computational Creativity researchers, and existing creative
  communities.  We should now add a fourth important ``stakeholder''
  group in computational creativity research: computer systems
  themselves.  System designers need to teach their systems how to
  make evaluations in way that is both reasonable and ethical.  This
  condition is exemplified by the preference for a ``non-zero sum''
  criterion for value introduced in Section \ref{sec:by-example}.
  \emph{A fourth challenge is for computational agents to evaluate
    their own creative process and products.}
\end{itemize}


%% A survey of word occurrences from a recent special issue of
%% \emph{Cognitive Computation} on ``Computational Creativity, Intelligence and Autonomy'' \cite{bishop-erden-special-issue} shows that related themes are broadly
%% active in the research community.  Here
%% \emph{italics} indicates that the word stem accounted for 0.1\% of the
%% article or more; added \textbf{\emph{bold}} indicates that it
%% accounted for 1\% or more.\footnote{Articles were converted to text
%%   via {\tt pdftotext -layout}, individual counts found via {\tt tr
%%     \textquotesingle~\textquotesingle~\textquotesingle\textbackslash
%%     n\textquotesingle~< file.txt | grep -c "stem*"}, and total word counts
%%   via {\tt wc -w}.  The corresponding counts for the \emph{current}
%%   paper are 12, \emph{25}, \emph{16}, \emph{44} and 12.7K.}

%% \medskip

%% \input{cog-comp-table}

%% \bigskip

%% Paper 4, Rob Saunders's \citeyear{saunders2012towards} ``Towards
%% Autonomous Creative Systems: A Computational Approach'' was the only
%% contributed paper to emphasise all four of our themes according to the
%% metric above.  Saunders asks: ``What would it mean to produce an
%% autonomous creative system? How might we approach this task? And, how
%% would we know if we had succeeded?''  He argues for an approach ``that
%% models personal motivations, social interactions and the evolution of
%% domains.''  Paper 10, d'Inverno and Luck's \citeyear{d2012creativity}
%% ``Creativity Through Autonomy and Interaction'', also contains a
%% theoretical engagement with these themes, and presents a formalism for
%% multi-agent systems that could usefully be adapted to model
%% serendipitous encounters.  Both papers are particularly concerned with
%% \emph{motivation}, a topic that relates to both the prepared mind and
%% the theme of embedded evaluation.

%% We believe that our clarifications to the multifaceted concept of
%% serendipity will help encourage future computer-aided (and
%% computer-driven) investigations of the above themes and their
%% interrelationships.  Our extension of SPECS to cover serendipity will
%% be useful for evaluating progress.  We discuss some of our related
%% research plans below.
