\section{Conclusion} \label{sec:conclusion}

%% In Section \ref{sec:literature-review}, we survey the broad literature
%% on serendipity, and examine prior applications of the concept of
%% serendipity in a computing context.
%
We began by surveying ``serendipity'', developing a broad historical
view, and several criteria which are computationally feasible.  Along
similar lines to \citeA{andre2009discovery}, we propose a two-phase
model.
%
%% In Section \ref{sec:background} we present our formal definition of
%% serendipity, drawing connections with historical examples and
%% presenting standards for evaluation.
%
Adapting the ``Standardised Procedure for Evaluating Creative
Systems'' we developed a set of assessment standards for serendipity.
%
%% Section \ref{sec:computational-serendipity} presents computational
%% case studies and thought experiments in terms of this model.
%
We used this model to examine several partial examples of serendipity,
and a thought experiment that exhibits all of the features of our
model.
%
%% Section \ref{sec:discussion} offers recommendations for researchers
%% working in computational creativity (a key research area concerned
%% with the computational modelling of serendipity), and describes our
%% own plans for future work.
We then extracted several corollaries of our definitions, which
outline a paradigm for serendipitous computing rooted in
\emph{autonomy}, \emph{learning}, \emph{sociality}, and \emph{embedded
  evaluation}.

%% Section \ref{sec:conclusion} reviews the argument and summarises the
%% limitations of our analysis.

% What answers have we offered?
The ideas presented in this article point to several possible
directions for implementation, and to further theoretical questions
about programming with design patterns.  In any case, considerable
concrete work remains to be done in order to realise our model in
code.  Even our hand-picked examples of prior art pale in comparison
to the examples of serendipitous discovery and invention from human
history.  It would seem that a fully-automated system that can
realistically be said to behave in a highly serendipitous manner has
not yet been built. \textbf{[Actually, that depends on what we say in
    the SPECS section, let's check.]}
% Further questions
Nevertheless, the theoretical work in this paper shows that it is
indeed possible to plan -- and program -- for serendipity.
