\section{Connecting our formal definition to literature} \label{sec:connections-to-formal-definition}

In this section, we give a short overview covering the etymology of
the term ``serendipity'' and trace its development in order to pin
down the key commonalities from many definitions and instances.  In
particular, we point out key conditions of serendipity, their
components and general characteristics, including environmental
factors.  The structure of this section follows and updates an earlier
survey from \citeA{pease2013discussion}, drawing connections with the
new formal model described above.

\subsubsection*{Key condition for serendipity}

\begin{itemize}
\item \textbf{Focus shift}: ``\emph{After removing several of the
  burdock burrs (seeds) that kept sticking to his clothes and his
  dog's fur,}~[de Mestral]~\emph{became curious as to how it
  worked. He examined them under a microscope, and noted hundreds of
  `hooks' that caught on anything with a loop, such as clothing,
  animal fur, or hair. He saw the possibility of binding two materials
  reversibly in a simple fashion, if he could figure out how to
  duplicate the hooks and loops.}''~\cite{wiki:velcro}
%
\inlineitem{This corresponds to the identification of $T^\star$, which
  is common to both sides of the diagram.  \citeA{creativity-crisis}
  write that: ``To be creative requires divergent thinking (generating
  many unique ideas) and then convergent thinking (combining those
  ideas into the best result).''  Accordingly $T^\star$ may be thought
  of as an evolving vector of interesting possibilities or ``strategic data'' \cite[p. 507]{merton1948bearing}.  In de
  Mestral's case, the initial idea of a hook-and-loop fastener
  occurred in 1941 -- followed by a full decade of experimentation
  before he was ready to file a patent claim.  }
\end{itemize}

\subsubsection*{Components of serendipity}

\begin{itemize}
\item \textbf{Prepared mind}: 
Fleming's ``prepared mind'' included his focus
on carrying out experiments to investigate influenza as well as his
previous experience that foreign substances in petri dishes can kill
bacteria.  He was concerned above all with the question ``Is there a
substance which is harmful to harmful bacteria but harmless to human
tissue?''  \cite[p. 161]{roberts}.
%%
%
\inlineitem{This corresponds to the prior
  training $p$ and $p^{\prime}$ in our diagram.}
\item \textbf{Serendipity trigger}: The trigger does not directly
  cause the outcome, but rather, inspires a new insight.  It was long
  known by Quechua medics that cinchona bark stops shivering.  In
  particular, it worked well to stop shivering in malaria patients, as
  was observed when malarial Europeans first arrived in Peru.  The
  joint appearance of shivering Europeans and a South American remedy
  was the trigger.  That an extract from cinchona bark can cure and
  can even prevent malaria was subsequently revealed.
%
\inlineitem{This corresponds to the stimulus $T$ in our diagram.}
%%
\item \textbf{Bridge}: These include reasoning techniques, such as
  abductive inference (what might cause a clear patch in a petri
  dish?); analogical reasoning (de Mestral constructed a target domain
  from the source domain of burs hooked onto fabric); and conceptual
  blending (Kekul\'e blended his knowledge of molecule structure with
  his vision of a snake biting its tail).  The bridge may also rely on
  new social arrangements, such as the formation of cross-cultural
  research networks.
%
\inlineitem{This corresponds to the actions based on $p^{\prime}$
  taken on $T^\star$ leading to $R$.}
%%
\item \textbf{Result}: This may be a new product, artefact, process,
  hypothesis, a new use for a material substance, and so on.  The
  outcome may contribute evidence in support of a known hypothesis, or
  a solution to a known problem.  Alternatively, the result may itself
  be a {\em new} hypothesis or problem.  The result may be a
  ``pseudoserendipitous'' in the sense that it was {\em sought}, while
  nevertheless arising from an unknown, unlikely, coincidental or
  unexpected source.  More classically, it is an \emph{unsought}
  finding, such as the discovery of the Rosetta stone.
%
\inlineitem{This corresponds to our $R$.  Note that $R$ may imply
  updates to $p$ or $p^{\prime}$ in further phases of research.}
\end{itemize}

\subsubsection*{Dimensions of serendipity}

Whereas the foregoing items are the central features of the
definition, the following further characterise the circumstances under
which serendipity occurs in practice.

\begin{itemize}
\item \textbf{Chance}: Fleming \citeyear{fleming} noted: ``There are
  thousands of different moulds'' -- and ``that chance put the mould
  in the right spot at the right time was like winning the Irish
  sweep.''
%
\inlineitem{One must assume that relatively few triggers $T^\star$
  that are identified as interesting actually lead to useful results;
  in other words, the process is fallible.}
%%
\item \textbf{Curiosity}: Venkatesh Rao \citeyear{rao2011tempo} refers
  to a \emph{cheap trick} that takes place early on in a narrative in
  order to establish the preliminary conditions of order.  Curiosity
  with can play this role, and can dispose a creative person to begin,
  or to continue, a search into unfamiliar territory.
%
\inlineitem{The prior training $p$ causes interesting features to be
  extracted, even if they are not necessarily useful; $p^{\prime}$
  asks how these features \emph{might} be useful.  }
%%
\item \textbf{Sagacity}: This old-fashioned word is related to
  ``wisdom,'' ``insight,'' and especially to ``taste'' -- and
  describes the attributes, or skill, of the discoverer that
  contribute to forming the bridge between the trigger and the result.
  In many cases, such as an entanglement with cockle-burs, many others
  will have already been in a similar position and not obtained an
  interesting result.  Once a phenomenon has been identified as
  interesting, the disposition of the investigator may lead to a
  dogged pursuit of a useful application or improvement.
%
\inlineitem{Rather than a simple look-up
  rule, $p^{\prime}$ involves creating new knowledge.}
%%
\item \textbf{Value}: Note that the chance ``discovery'' of, say, a
  \pounds 10 note may be seen as happy by the person who finds it,
  whereas the loss of the same note would generally be regarded as
  unhappy.  Positive judgements of serendipity by a third party would
  be less likely in scenarios in which ``One man's loss is another
  man's gain'' than in scenarios where ``One man's trash is another
  man's treasure.''  If possible we prefer this sort of independent
  judgement \cite{jordanous:12}.
%
\inlineitem{The evaluation $|R|>0$ may be carried out ``locally'' (as
  an embedded part of the process of invention of $R$) or ``globally''
  (i.e.~as an external process).  }
\end{itemize}

\subsubsection*{Environmental factors}

\begin{itemize}
\item \textbf{Dynamic world}: Information about the world develops
  over time, and is not presented as a complete, consistent whole.  In
  particular, value may come later.  Van Andel
  \citeyear[p. 643]{van1994anatomy} estimates that in twenty percent
  of innovations ``something was discovered before there was a demand
  for it.''
%
\inlineitem{$T$ (and $T^\star$) appears within a stream of data with
  indeterminacy.  There is a further feedback loop, insofar as
  products $R$ influence the future state.}
%%
\item \textbf{Multiple contexts}: One of the dynamical aspects at play
  may be the discoverer going back and forth between different
  contexts, with different stimuli.  3M employee Arthur Fry sang in a
  church choir and needed a good way to mark pages in his hymn book;
  he happened to have been attending seminars offered by his colleague
  Silver about restickable glue.
%
\inlineitem{This is reflected directly in our model by the difference
  between the ``context of discovery'' involving prior preparations
  $p$, and the ``context of invention'' involving prior preparations
  $p^{\prime}$.  Both of these may be subdivided further.}
%%
\item \textbf{Multiple tasks}: Even within what would typically be
  seen as a single context, a discoverer may take on multiple tasks
  that segment the context into sub-contexts, or that cause the
  investigator to look in more than one direction.  The tasks may have
  an interesting \emph{overlap}, or they may point to a \emph{gap} in
  knowledge.  As an example of the latter, Penzias and Wilson used a
  large antenna to detect radio waves that were relayed by bouncing
  off of satellites.  After they had removed interference effects due
  to radar, radio, and heat, they found residual ambient noise that
  couldn't be eliminated \cite{wiki:cosmic-radiation}.
%
\inlineitem{Both $T$ and $T^\star$ may be multiple, causing an
  individual process to fork into communicating sub-processes that
  involve different skills sets.}
%%
\item \textbf{Multiple influences}: The ``bridge'' from trigger to
  result is often found through a social network, thus, for instance
  Penzias and Wilson only understood the significance of their work
  after reading a preprint by Jim Peebles that hypothesised the
  possibility of measuring radiation released by the big bang
  \cite{wiki:cosmic-radiation}.
%
\inlineitem{The process as a whole may be multiplied out among
  different communicating investigators.}
\end{itemize}
