\subsubsection*{ Step 2: Evaluation standards for computational serendipity}
\begin{quote} {\em Using Step 1, clearly state what standards you use to evaluate the serendipity of your
    system. }\end{quote}

With our definition in mind, we propose the following standards for evaluating serendipity in computational systems:

%% Serendipity relies on a reassessment or reevaluation -- a \emph{focus shift} in which something that was previously uninteresting, of neutral, or even negative value, becomes interesting.

\begin{description}
\item[\emph{(\textbf{A - Definitional characteristics})}] \emph{The
  system can be said to have a \emph{\textbf{prepared mind}},
  consisting of previous experiences, background knowledge, a store of
  unsolved problems, skills, expectations, and (optionally) a current
  focus or goal.  It then processes a \emph{\textbf{serendipity
  trigger}} that is at least partially the result of factors outside
  of its control, including randomness or unexpected events.  The
  system then uses reasoning techniques and/or social or otherwise
  externally enacted alternatives to create a \emph{\textbf{bridge}}
  from the trigger to a result.  The \emph{\textbf{result}} is
  evaluated as useful, by the system and/or by an external source.}
\item[\emph{(\textbf{B - Dimensions})}] \emph{Serendipity, and its
  various dimensions, can be present to a greater or lesser degree.
  If the criteria above have been met, we generate ratings as
  estimated probabilities along several dimensions:
%
\emph{($\mathbf{a}$ - \textbf{chance})} how likely was this trigger to appear to
  the system?
%
\emph{($\mathbf{b}$ - \textbf{curiosity})} On a population basis, comparing
  similar circumstances, how likely was the trigger to be identified
  as interesting?
%
\emph{($\mathbf{c}$ - \textbf{sagacity})} On a population basis, comparing
  similar circumstances, how likely was it that the trigger
  would be turned into a result?
%
Finally, we ask, again, comparing similar results where possible:
\emph{($\mathbf{d}$ - \textbf{value})} How valuable is the result that
is ultimately produced?}
\begin{itemize}
\item \emph{Then $\mathbf{a}\times\mathbf{b}\times\mathbf{c}$ gives a
  likelihood score: low likelihood and high value is the criterion we use to say that the event was ``highly serendipitous.''}
\end{itemize}
\item[\emph{(\textbf{C - Factors})}] \emph{Finally, if the criteria
  from Part A are met, and if the event is deemed ``highly
  serendipitous'' according to the criteria in Part B, then in order
  to deepen our qualitative understanding of the serendipitous
  behaviour, we ask: To what extent does the system exist in a
  \emph{\textbf{dynamic world}}, spanning \emph{\textbf{multiple
      contexts}}, featuring \emph{\textbf{multiple tasks}}, and
  incorporating \emph{\textbf{multiple influences}}?}
\end{description}

\subsubsection*{Step 3: Testing our serendipitous system}

\begin{quote} {\em Test your serendipitous system against the standards stated in Step 2 and report the
results.}\end{quote}

We will develop several examples of the application of this framework
to examples from computing in Section
\ref{sec:computational-serendipity}.  First, a quick illustration
using one of the historical examples mentioned earlier.

\begin{mdframed}
\vspace{-.35cm}
\paragraph{De Mestral's invention of Velco\texttrademark}~\\[.05cm]
\begin{description}
\item[\emph{\textbf{prepared mind}}] De Mestral was a trained engineer and experienced inventor.
\item[\emph{\textbf{serendipity trigger}}] Burdock burs stuck to clothes and fur.
\item[\emph{\textbf{bridge}}] Viewing the burs under a microscope, followed by experiments to create a fastener that worked on the same principle.
\item[\emph{\textbf{result}}] Velcro\texttrademark
\item[\emph{\textbf{chance}}] High chance of encountering burs while out walking.
\item[\emph{\textbf{curiosity}}] Few people would investigate them under a microscope.
\item[\emph{\textbf{sagacity}}] Few people would conceive of a useful application or have the perseverance required to carry out product development.
\item[\emph{\textbf{value}}] High; hook-and-loop fasteners are widely used.
\begin{itemize}
\item Ruling: ``highly serendipitous''.
\end{itemize}
\item[\emph{\textbf{dynamic world}}] New fabrication principles required.
\item[\emph{\textbf{multiple contexts}}] Biomimicry (art imitating life).
\item[\emph{\textbf{multiple tasks}}] N/A.
\item[\emph{\textbf{multiple influences}}] N/A.
\end{description}
\end{mdframed}


%% In order to develop connections with our theoretical framework, and
%% because existing experiments have not been particularly strong, we
%% focus on a thought experiment in the
%% Section \ref{sec:computational-serendipity} detailing some of the
%% outcomes we would like to see, and some of the risks.
