\begin{thebibliography}{10}

\bibitem{alexander1979timeless}
{\sc Alexander, C.}
\newblock {\em The timeless way of building}.
\newblock Oxford University Press, 1979.

\bibitem{alexander1999origins}
{\sc Alexander, C.}
\newblock The origins of pattern theory: {T}he future of the theory, and the
  generation of a living world.
\newblock {\em Software, IEEE 16}, 5 (1999), 71--82.

\bibitem{alexander1977pattern}
{\sc Alexander, C., Ishikawa, S., and Silverstein, M.}
\newblock {\em {A} {P}attern {L}anguage: {T}owns, {B}uildings, {C}onstruction}.
\newblock Oxford University Press, 1977.

\bibitem{andersen2002dynamic}
{\sc Andersen, P.~B.}
\newblock Dynamic semiotics.
\newblock {\em Semiotica 139}, 1/4 (2002), 161--210.

\bibitem{andre2009discovery}
{\sc Andr{\'e}, P., Teevan, J., Dumais, S.~T., et~al.}
\newblock Discovery is never by chance: designing for (un) serendipity.
\newblock In {\em Proceedings of the seventh ACM conference on Creativity and
  cognition\/} (2009), ACM, pp.~305--314.

\bibitem{benfey}
{\sc Benfey, O.~T.}
\newblock August {K}ekul\'e and the {B}irth of the {S}tructural {T}heory of
  {O}rganic {C}hemistry in 1858.
\newblock {\em Journal of Chemical Education 35\/} (1958), 21--23.

\bibitem{bergin2012pedagogical}
{\sc Bergin, J., Eckstein, J., Volter, M., Sipos, M., Wallingford, E.,
  Marquardt, K., Chandler, J., Sharp, H., and Manns, M.~L.}
\newblock {\em {P}edagogical {P}atterns: {A}dvice for {E}ducators}.
\newblock Joseph Bergin Software Tools, 2012.

\bibitem{bergson2010creative}
{\sc Bergson, H.}
\newblock {\em {T}he {C}reative {M}ind}.
\newblock Greenwood Press, 1946 [1941].
\newblock (trans. Mabel L. Andison).

\bibitem{boden}
{\sc Boden, M.}
\newblock {\em {T}he {C}reative {M}ind: {M}yths and {M}echanisms}.
\newblock Weidenfield and Nicholson, London, 1990.

\bibitem{bouzy2001computer}
{\sc Bouzy, B., and Cazenave, T.}
\newblock {C}omputer {G}o: an {AI} oriented survey.
\newblock {\em Artificial Intelligence 132}, 1 (2001), 39--103.

\bibitem{boyd2004conversation}
{\sc Boyd, G.~M.}
\newblock Conversation theory.
\newblock In {\em Handbook of research for educational communications and
  technology}, D.~H. Jonassen, Ed., 2~ed. Lawrence Erlbaum, 2004, pp.~179--197.

\bibitem{brown1998antipatterns}
{\sc Brown, W.~J., McCormick, H.~W., Mowbray, T.~J., and Malveau, R.~C.}
\newblock {\em Anti{P}atterns: refactoring software, architectures, and
  projects in crisis}.
\newblock Wiley New York, 1998.

\bibitem{campbell}
{\sc Campbell, R.~L.}
\newblock On doing the impossible.
\newblock {\em Behavioural and Brain Sciences 17}, 3 (1994), 535 -- 537.
\newblock Open peer commentary on \cite{boden}.

\bibitem{Campos2002}
{\sc Campos, J., and Figueiredo, A.~D.}
\newblock {P}rogramming for {S}erendipity.
\newblock In {\em Proc. of the AAAI Fall Symposium on Chance Discovery -- The
  Discovery and Management of Chance Events\/} (2002).

\bibitem{companions-in-geography}
{\sc Cams, M.}
\newblock {`}{C}ompanions in {G}eography{'}: {T}he {S}ino-{E}uropean {E}ffort
  to {M}easure {C}hina, c.1685--1735.
\newblock {\em Imago Mundi 66}, 1 (2014), 136--137.

\bibitem{dialectique}
{\sc Chambers, E.}
\newblock Dialectique.
\newblock In {\em Encyclop{\'e}die ou Dictionnaire raisonn{\'e} des sciences,
  des arts et des m{\'e}tiers}, D.~Diderot and J.~le~Rond~d'Alembert, Eds.,
  vol.~4. Le Breton, David, Briasson, \& Durand, 1754.

\bibitem{charnley2014flowr}
{\sc Charnley, J., Colton, S., and Llano, M.~T.}
\newblock {T}he {F}lo{W}r framework: {A}utomated flowchart construction,
  optimisation and alteration for creative systems.
\newblock In {\em Proceedings of the 5th International Conference on
  Computational Creativity\/} (2014).

\bibitem{colton-flowcharting}
{\sc Colton, S., and Charnley, J.}
\newblock {T}owards a {F}lowcharting {S}ystem for {A}utomated {P}rocess
  {I}nvention.
\newblock In {\em Proceedings of the Fifth International Conference on
  Computational Creativity\/} (2014), D.~Ventura, S.~Colton, N.~Lavrac, and
  M.~Cook, Eds.

\bibitem{colton2014acid}
{\sc Colton, S., Cook, M., Hepworth, R., and Pease, A.}
\newblock On acid drops and teardrops: Observer issues in computational
  creativity.
\newblock In {\em Proceedings of the 7th {AISB} {S}ymposium on {C}omputing and
  {P}hilosophy\/} (2014).

\bibitem{stakeholder-groups-bookchapter}
{\sc Colton, S., Pease, A., Corneli, J., Cook, M., Hepworth, R., and Ventura,
  D.}
\newblock {S}takeholder {G}roups in {C}omputational {C}reativity {R}esearch and
  {P}ractice.
\newblock In {\em Computational Creativity Research: Towards Creative
  Machines}, T.~R. Besold, M.~Schorlemmer, and A.~Smaill, Eds., Thinking
  Machines: Studies in Computational Cognition. Atlantis - Springer,
  forthcoming.

\bibitem{colton-assessingprogress}
{\sc Colton, S., Pease, A., Corneli, J., Cook, M., and Llano, T.}
\newblock {A}ssessing {P}rogress in {B}uilding {A}utonomously {C}reative
  {S}ystems.
\newblock In {\em Proceedings of the Fifth International Conference on
  Computational Creativity\/} (2014), D.~Ventura, S.~Colton, N.~Lavrac, and
  M.~Cook, Eds.

\bibitem{coplien1997pattern}
{\sc Coplien, J.~O., and Woolf, B.}
\newblock A pattern language for writers' workshops.
\newblock {\em {C++} report 9\/} (1997), 51--60.

\bibitem{de2013turing}
{\sc de~Sojo, A.~A., Ares, J., Lara, J.~A., Lizcano, D., Mart{\'\i}nez, M.~A.,
  and Pazos, J.}
\newblock {T}uring and the {S}erendipitous {D}iscovery of the {M}odern
  {C}omputer.
\newblock {\em Foundations of Science\/} (2013), 1--13.

\bibitem{deacon2006emergence}
{\sc Deacon, T.~W.}
\newblock Emergence: {T}he hole at the wheel{'}s hub.
\newblock In {\em The re-emergence of emergence}, P.~Clayton and P.~Davies,
  Eds. Oxford University Press, 2006, pp.~111--150.

\bibitem{delanda1993virtual}
{\sc Delanda, M.}
\newblock Virtual environments and the emergence of synthetic reason.
\newblock {\em South Atlantic Quarterly 92}, 4 (1993), 793--815.

\bibitem{delanda2005deleuze}
{\sc DeLanda, M.}
\newblock {D}eleuze and the {O}pen {E}nded {B}ecoming of the {W}orld.
\newblock In {\em {C}haos/{C}ontrol: {C}omplexity {C}onference, {U}niversity of
  {B}ielefeld, {G}ermany\/} (1998).

\bibitem{delanda2011philosophy}
{\sc DeLanda, M.}
\newblock {\em Philosophy and simulation: the emergence of synthetic reason}.
\newblock Continuum, 2011.

\bibitem{deleuze1991bergsonism}
{\sc Deleuze, G.}
\newblock {\em Bergsonism}.
\newblock Zone, 1988 [1966].
\newblock (trans. Hugh Tomlinson and Barbara Habberjam).

\bibitem{deleuze1994difference}
{\sc Deleuze, G.}
\newblock {\em Difference and repetition}.
\newblock Bloomsbury Academic, 2004 [1968].
\newblock (trans. Paul Patton).

\bibitem{dewey1997we}
{\sc Dewey, J.}
\newblock {\em How we think}.
\newblock Courier Dover Publications, 1997 [1910].

\bibitem{chumaceiro1995serendipity}
{\sc D{\'{\i}}az~de Chumaceiro, C.~L.}
\newblock Serendipity or pseudoserendipity? {U}nexpected versus desired
  results.
\newblock {\em The Journal of Creative Behavior 29}, 2 (1995), 143--147.

\bibitem{Donoghue2002}
{\sc Donoghue, D., and Crean, B.}
\newblock {Searching for Serendipitous Analogies}.
\newblock In {\em European Conference on Artifical Intelligence (ECAI),
  Workshop on Creative Systems\/} (July, 2002).

\bibitem{Donoghue2012}
{\sc Donoghue, D., and Keane, M.}
\newblock {A Creative Analogy Machine: Results and Challenges Analogical
  Retrieval Phase-Model}.
\newblock In {\em Proceedings of the International Conference on Computational
  Creativity 2012\/} (2012), pp.~17--24.

\bibitem{eco2013serendipities}
{\sc Eco, U.}
\newblock {\em Serendipities: {L}anguage and lunacy}.
\newblock Columbia University Press, 2013.

\bibitem{Figueiredo2001}
{\sc Figueiredo, A.~D., and Campos, J.}
\newblock {The Serendipity Equations}.
\newblock In {\em Proc. of ICCBR-4\/} (2001), R.~Weber and C.~G. von
  Wangenheim, Eds.

\bibitem{tce-postits}
{\sc Flavell-While, C.}
\newblock {S}pencer {S}ilver and {A}rthur {F}ry: the chemist and the tinkerer
  who created the {P}ost-it {N}ote.
\newblock {\em The Chemical Engineer\/} (August 2012), 53--55.

\bibitem{fleming}
{\sc Fleming, A.}
\newblock Penicillin.
\newblock In {\em Nobel Lectures, Physiology or Medicine, 1942-1962}. Elsevier,
  1964.

\bibitem{forth2013serena}
{\sc Forth, J., Giannimaras, T., Wiggins, G.~A., Stewart, R., Bental, D.,
  Aylett, R., Maxwell, D., Mehrpouya, H., Shek, J., and Woods, M.}
\newblock {SerenA}: A multi-site pervasive agent environment that supports
  serendipitous discovery in research.
\newblock In {\em Advances on Practical Applications of Agents and Multi-Agent
  Systems}. Springer, 2013, pp.~85--96.

\bibitem{foster2003serendipity}
{\sc Foster, A., and Ford, N.}
\newblock Serendipity and information seeking: an empirical study.
\newblock {\em Journal of Documentation 59}, 3 (2003), 321--340.

\bibitem{gabriel1996patterns}
{\sc Gabriel, R.~P.}
\newblock {\em {P}atterns of {S}oftware}.
\newblock Oxford University Press New York, 1996.

\bibitem{gabriel2002writer}
{\sc Gabriel, R.~P.}
\newblock {\em {W}riter's {W}orkshops and the {W}ork of {M}aking {T}hings}.
\newblock Addison-Wesley Longman Publishing Co., Inc., 2002.

\bibitem{geertz1973interpretation}
{\sc Geertz, C.}
\newblock {\em The interpretation of cultures: {S}elected essays}.
\newblock Basic Books (AZ), 1973.

\bibitem{goguen1999introduction}
{\sc Goguen, J.}
\newblock An introduction to algebraic semiotics, with application to user
  interface design.
\newblock In {\em Computation for metaphors, analogy, and agents}. Springer,
  1999, pp.~242--291.

\bibitem{goodyear1855gum}
{\sc Goodyear, C.}
\newblock {\em Gum-Elastic and its Varieties, with a Detailed Account of its
  Applications and Uses, and of the Discovery of Vulcanization}.
\newblock Pub. for the author, 1855.

\bibitem{hofstadter1994copycat}
{\sc Hofstadter, D.~R., and Mitchell, M.}
\newblock The copycat project: {A} model of mental fluidity and analogy-making.
\newblock In {\em {F}luid {C}oncepts and {C}reative {A}nalogies: {C}omputer
  {M}odels of the {F}undamental {M}echanisms}, D.~R. Hofstadter and the Fluid
  Analogies Research~Group, Eds. Basic Books, 1994, pp.~205--267.

\bibitem{Iaquinta2008}
{\sc Iaquinta, L., Gemmis, M., Lops, P., Semeraro, G., Filannino, M., and
  Molino, P.}
\newblock {I}ntroducing {S}erendipity in a {C}ontent-{B}ased {R}ecommender
  {S}ystem.
\newblock {\em 8th Int. Conf. on Hybrid Intelligent Systems\/} (Sept. 2008),
  168--173.

\bibitem{imbens1994identification}
{\sc Imbens, G.~W., and Angrist, J.~D.}
\newblock Identification and estimation of local average treatment effects.
\newblock {\em Econometrica: Journal of the Econometric Society\/} (1994),
  467--475.

\bibitem{jevons:1877}
{\sc Jevons, W.~S.}
\newblock {\em {P}rinciples of {S}cience: {A} treatise on logic and scientific
  method}.
\newblock Macmillan, London, {1913 [1874]}.

\bibitem{jin1975art}
{\sc Jin'Ichi, K., Brazell, K., and Cook, L.}
\newblock The {A}rt of {R}enga.
\newblock {\em Journal of Japanese Studies\/} (1975), 29--61.

\bibitem{jordanous:12}
{\sc Jordanous, A.}
\newblock {A} {S}tandardised {P}rocedure for {E}valuating {C}reative {S}ystems:
  {C}omputational {C}reativity {E}valuation {B}ased on {W}hat it is to be
  {C}reative.
\newblock {\em Cognitive Computation 4}, 3 (2012), 246--279.

\bibitem{jordanous12jims}
{\sc Jordanous, A., and Keller, B.}
\newblock What makes musical improvisation creative?
\newblock {\em Journal of Interdisciplinary Music Studies 6}, 2 (2012),
  151--175.

\bibitem{kohls2010structure}
{\sc Kohls, C.}
\newblock The structure of patterns.
\newblock In {\em {P}roceedings of the 17th {C}onference on {P}attern
  {L}anguages of {P}rograms\/} (2010), ACM, p.~12.

\bibitem{kohls2011structure}
{\sc Kohls, C.}
\newblock The structure of patterns: part {II}--qualities.
\newblock In {\em {P}roceedings of the 18th {C}onference on {P}attern
  {L}anguages of {P}rograms\/} (2011), ACM, p.~27.

\bibitem{lenat1983eurisko}
{\sc Lenat, D.~B.}
\newblock {EURISKO}: a program that learns new heuristics and domain concepts:
  the nature of heuristics {III}: program design and results.
\newblock {\em Artificial Intelligence 21}, 1 (1983), 61--98.

\bibitem{Makri2012}
{\sc Makri, S., and Blandford, A.}
\newblock Coming across information serendipitously - {P}art 1: A process
  model.
\newblock {\em Journal of Documentation 68\/} (2012), 684--705.

\bibitem{Makri2012a}
{\sc Makri, S., and Blandford, A.}
\newblock Coming across information serendipitously - {P}art 2: A
  classification framework.
\newblock {\em Journal of Documentation 68\/} (2012), 706--724.

\bibitem{DBLP:journals/jetai/Marshall06}
{\sc Marshall, J.~B.}
\newblock A self-watching model of analogy-making and perception.
\newblock {\em J. Exp. Theor. Artif. Intell. 18}, 3 (2006), 267--307.

\bibitem{maxwell2012designing}
{\sc Maxwell, D., Woods, M., Makri, S., Bental, D., Kefalidou, G., and
  Sharples, S.}
\newblock Designing a semantic sketchbook to create opportunities for
  serendipity.
\newblock In {\em Proceedings of the 26th Annual BCS Interaction Specialist
  Group Conference on People and Computers\/} (2012), British Computer Society,
  pp.~357--362.

\bibitem{mead1932philosophy}
{\sc Mead, G.~H.}
\newblock {\em The philosophy of the present}.
\newblock Prometheus Books, 1932.

\bibitem{dewey-by-mead}
{\sc Mead, G.~H.}
\newblock The philosophy of {J}ohn {D}ewey.
\newblock {\em International Journal of Ethics 46}, 1 (1935), 64--81.

\bibitem{merton}
{\sc Merton, R.~K., and Barber, E.}
\newblock {\em {T}he {T}ravels and {A}dventures of {S}erendipity: {A} study in
  {S}ociological {S}emantics and the {S}ociology of {S}cience}.
\newblock Princeton University Press, New Jersey, USA, 2004.

\bibitem{minsky1967programming}
{\sc Minsky, M.}
\newblock Why programming is a good medium for expressing poorly understood and
  sloppily formulated ideas.
\newblock {\em Design and Planning II-Computers in Design and Communication\/}
  (1967), 120--125.

\bibitem{mitchell1993analogy}
{\sc Mitchell, M.}
\newblock {\em {A}nalogy-making as perception: {A} computer model}.
\newblock MIT Press, 1993.

\bibitem{conversation-theory-review}
{\sc Pask, G.}
\newblock Review of conversation theory and a protologic (or protolanguage),
  lp.
\newblock {\em ECTJ 32}, 1 (1984), 3--40.

\bibitem{pearce-boden-and-beyond}
{\sc Pearce, M.}
\newblock {B}oden and {B}eyond: \emph{{T}he {C}reative {M}ind} and its
  {R}eception in the {A}cademic {C}ommunity.
\newblock Tech. rep., City University, 2010.

\bibitem{pease2013discussion}
{\sc Pease, A., Colton, S., Ramezani, R., Charnley, J., and Reed, K.}
\newblock {A} {D}iscussion on {S}erendipity in {C}reative {S}ystems.
\newblock In {\em Proceedings of the Fourth International Conference on
  Computational Creativity\/} (2013).

\bibitem{poincare1910creation}
{\sc Poincar{\'e}, H.}
\newblock Mathematical creation.
\newblock {\em The Monist 20}, 3 (1910), 321--335.

\bibitem{poincare2013science}
{\sc Poincar{\'e}, H.}
\newblock {\em Science and method}.
\newblock Courier Dover Publications, 2013 [1914].
\newblock (trans. Francis Maitland).

\bibitem{eloge-de-la-simulation}
{\sc Qu{\'e}au, P.}
\newblock {\em {\'E}loge de la simulation: de la vie des langages {\`a} la
  synth{\`e}se des images}.
\newblock Champ Vallon/INA, 1986.

\bibitem{rao2011tempo}
{\sc Rao, V.}
\newblock {\em Tempo: {T}iming, tactics and strategy in narrative-driven
  decision-making}.
\newblock Ribbonfarm, Inc., 2011.

\bibitem{reichardt1969cybernetic}
{\sc Reichardt, J.}
\newblock {\em Cybernetic serendipity: the computer and the arts}.
\newblock Praeger, 1969.

\bibitem{riedel2005edward}
{\sc Riedel, S.}
\newblock Edward {J}enner and the history of smallpox and vaccination.
\newblock {\em Proceedings (Baylor University Medical Center) 18}, 1 (2005),
  21.

\bibitem{roberts}
{\sc Roberts, R.~M.}
\newblock {\em {S}erendipity: {A}ccidental {D}iscoveries in {S}cience}.
\newblock John Wiley and Sons, Inc., USA, 1989.

\bibitem{rodgers2002defining}
{\sc Rodgers, C.}
\newblock {D}efining reflection: {A}nother look at {J}ohn {D}ewey and
  reflective thinking.
\newblock {\em The Teachers College Record 104}, 4 (2002), 842--866.

\bibitem{rowe1994creativity}
{\sc Rowe, J.}
\newblock Creativity: {M}etarules and emergent systems.
\newblock {\em Behavioral and Brain Sciences 17}, 03 (1994), 550--551.
\newblock Open peer commentary on \cite{boden}.

\bibitem{schuler2008liberating}
{\sc Schuler, D.}
\newblock {\em Liberating voices: {A} pattern language for communication
  revolution}.
\newblock MIT Press, 2008.

\bibitem{singh2004big}
{\sc Singh, S.}
\newblock {\em Big bang}.
\newblock Rizzoli, 2004.

\bibitem{turing-intelligent}
{\sc Turing, A.}
\newblock Intelligent machinery, a heretical theory, 1951.
\newblock A lecture given to {`}51 Society{'} at Manchester.

\bibitem{van1994anatomy}
{\sc Van~Andel, P.}
\newblock {A}natomy of the {U}nsought {F}inding. {S}erendipity: {O}rgin,
  {H}istory, {D}omains, {T}raditions, {A}ppearances, {P}atterns and
  {P}rogrammability.
\newblock {\em The British Journal for the Philosophy of Science 45}, 2 (1994),
  631--648.

\bibitem{von2003essays}
{\sc von Foerster, H.}
\newblock {\em {U}nderstanding {U}nderstanding: {E}ssays on {C}ybernetics and
  {C}ognition}.
\newblock Springer, 2003.

\bibitem{wallas1926art}
{\sc Wallas, G.}
\newblock {\em The art of thought}.
\newblock J. Cape, 1926.

\bibitem{wiki:cosmic-radiation}
{\sc Wikipedia}.
\newblock Discovery of cosmic microwave background radiation --- {W}ikipedia{,}
  {T}he {F}ree {E}ncyclopedia, 2014.

\bibitem{wiki:velcro}
{\sc Wikipedia}.
\newblock George de {M}estral --- {W}ikipedia{,} {T}he {F}ree {E}ncyclopedia,
  2014.

\bibitem{yuan2008towards}
{\sc Yuan, T., Schulze, J., Devereux, J., and Reed, C.}
\newblock Towards an arguing agents competition: Building on argumento.
\newblock In {\em Proceedings of IJCAI'2008 Workshop on Computational Models of
  Natural Argument\/} (2008).

\end{thebibliography}
